---
title: "Customizing ggplot"
description: |
  A new article created using the Distill format.
author:
  - name: Sean Conway
date: "`r Sys.Date()`"
output: distill::distill_article
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## intro  

This script is meant to provide some demonstrations of how to customize the appearance of a finalized `ggplot` object.  

## libraries  

```{r warning=F, message=F}
library(dplyr)
library(ggplot2)
library(gapminder)
```

## data  
  
In this script, we'll be working with the `gapminder` dataset from the package of the same name.  

```{r}
data("gapminder")
gapminder %>%
  rmarkdown::paged_table()
```

## analysis
  
The main purpose of this analysis is to assess how the median life expectancy has changed over time on each continent. 

```{r}
cont <- gapminder %>%
  mutate(continent=as.factor(continent)) %>%
  group_by(continent,year) %>%
  summarise(med=median(lifeExp,na.rm=T)) %>%
  ungroup()
rmarkdown::paged_table(cont)
```

## plotting  
  
### Basic plot   


```{r}
pl <- ggplot(cont, aes(year,med,col=continent))+
  geom_path()
pl 
```
  
### Adding text labels using `ggrepel`, other customizations
  
The `ggrepel` package allows us to plot text labels while making sure they do not overlap.  

We also use `labs()` to change plot labels, and the `scale_*_continuous()` functions to modify axis ticks and breaks.  

```{r warning=F}
pl_lab <- pl + 
  ggrepel::geom_label_repel(aes(label=ifelse(year==min(year),as.character(continent),NA_character_)))+
  labs(x="Year",
       y="Median Life Expectancy",
       title="Median Life Expectancy Over Time",
       subtitle="Split by Continent")+
  scale_x_continuous(limits=c(1950,2010),breaks=seq(1950,2010,10))+
  scale_y_continuous(limits=c(30,90),breaks = seq(30,90,10))
pl_lab
```

### More customizations using `theme()`  

The `theme()` function is the workhorse function for customizing the appearance of a `ggplot` object. The `element_text()` functions allows us to change the appearance of the text for various aspects of the plot. 

```{r}
pl_lab_1 <- pl_lab + 
  theme(plot.title = element_text(hjust=0.5,size=14),
        plot.subtitle = element_text(hjust=0.5,size=12),
        axis.text = element_text(size=10))
pl_lab_1
```
  
Here, we use `element_rect() to change the background color and `element_line()` to modify gridlines.  

```{r}
pl_lab_1+
  theme(panel.background = element_rect(fill="white"),
        panel.grid = element_line(color="black",size=.1))
```

### Customizations using `ggthemes`  

`ggthemes` is a great `R` package that has ready-made themes that can change the appearance of our plot.

```{r}
pl_lab_1 + 
  ggthemes::theme_base()
```
  
  
```{r}
pl_lab_1 +
  ggthemes::theme_fivethirtyeight()
```


```{r}
pl_lab_1+
  ggthemes::theme_excel()
```

```{r}
pl_lab_1+
  ggthemes::theme_few()
```

## Final Plot  

This code chunk is the full set of code for making the finalized `ggplot` object.  

```{r}
final_pl <- ggplot(cont, aes(year,med,col=continent))+
  geom_path()+ 
  ggrepel::geom_label_repel(aes(label=ifelse(year==min(year),as.character(continent),NA_character_)))+
  labs(x="Year",
       y="Median Life Expectancy",
       title="Median Life Expectancy Over Time",
       subtitle="Split by Continent")+
  scale_x_continuous(limits=c(1950,2010),breaks=seq(1950,2010,10))+
  scale_y_continuous(limits=c(30,90),breaks = seq(30,90,10))+
  ggthemes::theme_few()+
  theme(plot.title = element_text(hjust=0.5,size=14),
        plot.subtitle = element_text(hjust=0.5,size=12),
        axis.text = element_text(size=10))
final_pl
```









